Towards a Verified Tableau Prover for a Ground Fragment of Set Theory

### 2023-02-17 paper/CADE/paper.tex
Continue with paper

\newcommand{\fmAnd}[2]{#1 $\boldsymbol{\land}$ #2}
\newcommand{\fmOr}[2]{#1 $\boldsymbol{\lor}$ #2}
\newcommand{\fmNegSymbol}{\boldsymbol{\neg}}
\newcommand{\fmNeg}[1]{$\fmNegSymbol\:$#1}
\newcommand{\fmAtom}{\textbf{A}}

@@

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  \keywords{First TODO \and Second TODO}
\end{abstract}

@@

@@

\subsection{Contributions}
In this paper, we present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation follows the paper closely but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
  \item We restrict one branching rule TODO.
\end{itemize}

@@

While a lot of work has gone into the verification of decision procedures, seemingly little has been put towards the verification of tableau calculi.
In Lean, there is a verified tableau calculus for basic modal logic~\cite{modal_logic_lean}.
In Coq, there is a formalisation of a tableau calculus for ecumenical logic~\cite{ecumenical_logic}; it is, however, missing a completeness proof.
Concerning Isabelle/HOL, there is formalisation of a sound and complete tableau calculus for hybrid logic~\cite{hybrid_logic} available in the \textit{Archive of Formal Proofs}~\cite{hybrid_logic_afp}.
The termination proof is part of ongoing work~\cite{hybrid_logic_workshop}.
TODO: Mention SteP, Check reviewer comment.
% TODO: Formalisation de la logique temporelle dans le coq.

@@

Similarly to \lstinline!vars!, we introduce the overloaded constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!, which returns all the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we consider the subformulae of a formula and define a function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes them.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the usual way (cf. Appendix TODO).

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation as displayed in TODO uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises for which arguments the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \autoref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition.

@@

In particular, we will use this notion in \autoref{sec:bound} where we derive an upper bound for the cardinality of well-formed branches.
This is then used in \autoref{sec:correct} to justify the termination of the decision procedure. 
Before we come to that, though, we prove soundness and completeness in Section~\ref{sec:sound} and~\ref{sec:complete}, respectively.
In \autoref{sec:correct}, we also show that both properties easily transfer to \lstinline!mlss_proc! which, together with termination, establishes that it is a decision procedure.

@@

The proof of \hyperref[lst:lemma_2]{\lstinline!lemma_2!} is more technical than interesting so we refer the reader to the formalisation or paper TODO.

@@

\begin{lstlisting}[belowskip=0pt,label={lst:bclosed_sound}]
lemma bclosed_sound:
  assumes bclosed b shows $\exists$$\phi$ $\in$ set b. M $\not\models$ $\phi$
\end{lstlisting}
\begin{proof}
  It is clear that, for any \lstinline!s!, neither does \lstinline!M! model \lstinline!s $\in$ $\emptyset$! nor \lstinline!s $\neqS$ s!.
  Furthermore, no model can satisfy both \lstinline!$\phi$! and \lstinline!&\fmNeg{$\phi$}&! at the same time.
  Lastly, a membership cycle is impossible since the membership relation of \lstinline!hf! is well-founded (see \autoref{sec:semantics} TODO).
\end{proof}

@@

\section{Bounding the Cardinality of a Branch\label{sec:bound}}
All in all, we get the following lemma.
\begin{lstlisting}[label={lst:card_branch}]
lemma card_wf_branch_ub:
  assumes wf_branch b
  shows |set b|
      $\leq$ 2 * |subfms (last b)| + 16 * |subterms (last b)|$^\text{4}$
\end{lstlisting}

\section{Total Correctness of the Decision Procedure\label{sec:correct}}
We first prove the termination of the procedure for well-formed branches, i.e.\ every well-formed branch is in the domain of \lstinline!mlss_proc_branch!.
\begin{lstlisting}[belowskip=0pt]
lemma mlss_proc_branch_dom_if_wf_branch:
  assumes wf_branch b shows mlss_proc_branch_dom b
\end{lstlisting}
\begin{proof}
  We proceed by strong induction on the difference between the upper bound due to \hyperref[lst:card_branch]{\lstinline!card_wf_branch_ub!} and \lstinline!|b|!.

  We only have one inductive case where the induction hypothesis holds for all well-formed branches \lstinline!b'! where the difference between the upper bound and the cardinality is smaller than that of \lstinline!b!.
  We strengthen the assumptions of the induction hypothesis to obtain
\begin{lstlisting}
  $\llbracket$ &\expandss{b'}{b}&; set b $\subset$ set b' $\rrbracket$ $\Longrightarrow$ mlss_proc_branch_dom b'
\end{lstlisting}
for all \lstinline!b'!.
The former strengthening is admissible since, by definition, every expansion of a well-formed branch is again well-formed while the latter ensures that the cardinality of \lstinline!b'! is larger than the cardinality of \lstinline!b!, implying that the difference to the upper bound decreases.

Now, consider the case where \lstinline!b! is saturated.
Then, the procedure terminates without recursive calls and we immediately get \lstinline!mlss_proc_branch_dom b! using the corresponding introduction rule for the domain predicate.

If \lstinline!b! is not saturated, we can expand \lstinline!b! using a linear or a branching rule, which adds new formulae to the resulting branch(es).
In the first case, we can thus choose a branch \lstinline!b'! with \lstinline!SOME! that linearly expands \lstinline!b!.
Moreover, the domain predicate holds for \lstinline!b' @ b! since the assumptions of the modified induction hypothesis are fulfilled.
Ultimately, we use the corresponding introduction rule
\begin{lstlisting}
  $\llbracket$ $\neg$ lin_sat b; mlss_proc_branch_dom (b' @ b) $\rrbracket$
  $\Longrightarrow$ mlss_proc_branch_dom b
\end{lstlisting}
for the domain predicate to obtain \lstinline!mlss_proc_branch b!.

The case of a branching rule is analogous.
\end{proof}

\noindent The above lemma allows us to use the computation induction rule of \lstinline!mlss_proc_branch! on well-formed branches, which we use to prove soundness and completeness.
As both proofs are essentially an application of soundness, respectively completeness, of the calculus, we refer the reader to the formalisation.
\begin{lstlisting}
lemma mlss_proc_branch_complete:
  fixes b :: 'a branch
  assumes wf_branch b and $\neg$ mlss_proc_branch b
  assumes infinite (UNIV :: 'a set)
  shows $\exists$M. M $\models$ last b

@@

@@

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule{1-1} \cmidrule{3-3}
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}

@@

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules.\label{tab:bexpands} We write \lstinline!$\phi$! for \lstinline!last b! here.}
\end{table*}

>>>>

\newcommand{\fmAndSymbol}{\boldsymbol{\land}}
\newcommand{\fmAnd}[2]{#1 $\fmAndSymbol$ #2}
\newcommand{\fmOr}[2]{#1 $\boldsymbol{\lor}$ #2}
\newcommand{\fmNegSymbol}{\boldsymbol{\neg}}
\newcommand{\fmNeg}[1]{$\fmNegSymbol\:$#1}
\newcommand{\fmAtom}{\textbf{A}}

@@

\begin{abstract}
  Using Isabelle/HOL, we verify the state-of-the-art decision procedure for multi-level syllogistic with singleton (\MLSS{} for short), which is a ground fragment of set theory.
  We formalise the syntax and semantics of \MLSS{} as well as a sound and complete tableau calculus for it.
  We also provide an abstract specification of a decision procedure that applies the rules of the calculus exhaustively and prove its termination.
  \keywords{First TODO \and Second TODO\todo{keywords}}
\end{abstract}

@@

\clearpage

@@

\subsection{Contributions}
In this paper, we present a formalisation in Isabelle/HOL of a tableau calculus for \MLSS{} due to \citeauthor{new_fast_tableau}~\cite{new_fast_tableau}\cite[Chapter 14]{set_theory}.
We prove soundness and completeness of the calculus and give an abstract specification of a decision procedure that applies the rules of the calculus exhaustively.
To obtain total correctness of the procedure, we prove its termination.
The formalisation follows the paper closely but gives a more thorough account of some important details:
\begin{itemize}
  \item We deliver the omitted proof of Lemma~2 in the paper~\cite{new_fast_tableau}, which is a key building block for the completeness proof of the calculus.
  \item The formal proof of completeness lead to the discovery that the calculus was missing a rule for eliminating double negation. 
  \item We derive an explicit upper bound for the number of formulae in a branch of the tableau.
  \item We restrict one branching rule\todo{Elaborate}.
\end{itemize}

@@

While a lot of work has gone into the verification of decision procedures, seemingly little has been put towards the verification of tableau calculi.
In Lean, there is a verified tableau calculus for basic modal logic~\cite{modal_logic_lean}.
In Coq, there is a formalisation of a tableau calculus for ecumenical logic~\cite{ecumenical_logic}; it is, however, missing a completeness proof.
Concerning Isabelle/HOL, there is formalisation of a sound and complete tableau calculus for hybrid logic~\cite{hybrid_logic} available in the \textit{Archive of Formal Proofs}~\cite{hybrid_logic_afp}.
The termination proof is part of ongoing work~\cite{hybrid_logic_workshop}.
\todo[inline]{Mention SteP, Check reviewr comment}
% TODO: Formalisation de la logique temporelle dans le coq.

@@

Similarly to \lstinline!vars!, we introduce the overloaded constant \lstinline!subterms :: 'b $\Rightarrow$ 'a pset_term set!, which returns all the set terms that are subterms of a set term, set atom, or formula, respectively. Lastly, we consider the subformulae of a formula and define a function \lstinline!subfms :: 'a fm $\Rightarrow$ 'a fm set! that computes them.
The functions \lstinline!subterms! and \lstinline!subfms! are defined in the usual way (cf. Appendix \todo{reference Appendix}).

@@

We abstractly specify the behaviour of the decision procedure, leaving an executable specification from which code can be generated to future work.
The implementation as displayed in \todo{How to reference algorithm?} uses a couple of features of Isabelle/HOL's function package:
instead of defining the function via pattern matching, we specify the equations of the function as conditional rewrite rules.
This requires us to prove that the assumptions of the equations are non-overlapping, which is done by automation.
The other concern is that Isabelle/HOL requires functions to be total, so a recursive function needs to terminate in order for it to be well-defined;
nevertheless, the termination proof is separated from the definition of the function for modularity.
The function package maintains the soundness of the definition by introducing a so-called domain predicate \lstinline!mlss_proc_branch_dom! which characterises for which arguments the function terminates.
Each equation of the function is guarded by an assumption that the predicate holds for the argument.
In \autoref{sec:correct}, we will show that the domain predicate holds for the context that \lstinline!mlss_proc_branch! is called in.
Before we go into more detail on how the termination is proved, we discuss its definition.

@@

In particular, we will use this notion in \autoref{sec:correct} to state an upper bound for the cardinality of well-formed branches, with which we justify the termination of the decision procedure. 
Before we come to that, though, we prove soundness and completeness in Section~\ref{sec:sound} and~\ref{sec:complete}, respectively.
In \autoref{sec:correct}, we also show that both properties easily transfer to \lstinline!mlss_proc! which, together with termination, establishes that it is a decision procedure.

@@

The proof of \hyperref[lst:lemma_2]{\lstinline!lemma_2!} is more technical than interesting so we refer the reader to the formalisation.

@@

\begin{lstlisting}[belowskip=0pt,label={lst:bclosed_sound}]
lemma bclosed_sound:
  assumes bclosed b shows $\exists$$\phi$ $\in$ set b. M $\not\models$ $\phi$
\end{lstlisting}
\begin{proof}
  It is clear that, for any \lstinline!s!, neither does \lstinline!M! model \lstinline!s $\in$ $\emptyset$! nor \lstinline!s $\neqS$ s!.
  Furthermore, no model can satisfy both \lstinline!$\phi$! and \lstinline!&\fmNeg{$\phi$}&! at the same time.
  Lastly, a membership cycle is impossible since the membership relation of \lstinline!hf! is well-founded.
\end{proof}

@@

\section{Total Correctness of the Decision Procedure\label{sec:correct}}
We first demonstrate the termination of the procedure for well-formed branches, i.e.\ every well-formed branch is in the domain of \lstinline!mlss_proc_branch!.
To this end, we derive an upper bound for the number of distinct formulae in a branch whose proof we omit here for brevity (cf. Appendix\todo{Reference Appendix}).
We should point out that this bound is not to be construed as the complexity of the procedure as it may create exponentially many branches in general.
\begin{lstlisting}[label={lst:card_branch}]
lemma card_wf_branch_ub:
  assumes wf_branch b
  shows |set b| $\leq$ 2$\:$*$\,$|subfms (last b)| + 16$\:$*$\,$|subterms (last b)|$^\text{4}$
\end{lstlisting}
Remember that \lstinline!mlss_proc_branch! only applies a linear expansion rule to a branch if the application results in new formulae.   
Moreover, the subsumption conditions of the branching expansion rules ensure that each of the newly created branches contain new formulae.
Ultimately, we conclude that the procedure must terminate because the number of formualae increases in each step but it is also bounded. 
\begin{lstlisting}
lemma mlss_proc_branch_dom_if_wf_branch:
  assumes wf_branch b shows mlss_proc_branch_dom b
\end{lstlisting}

The above lemma allows us to use the computation induction rule of \lstinline!mlss_proc_branch! on well-formed branches, which we use to prove soundness and completeness.
As both proofs are essentially an application of soundness, respectively completeness, of the calculus, we refer the reader to the formalisation.
\begin{lstlisting}
lemma mlss_proc_branch_complete:
  fixes b :: 'a branch
  assumes wf_branch b and $\neg$ mlss_proc_branch b
  assumes infinite (UNIV :: 'a set)
  shows $\exists$M. M $\models$ last b

@@

\section{Dealing with Urelements}
In the introduction, we motivated the verification of \lstinline!mlss_proc! with the prospect of integrating it as a tactic into Isabelle.
For this to work, we need to replay a proof in the tableau calculus through Isabelle/HOL, i.e.\ we need to map every branch expansion rule to a corresponding theorem in Isabelle/HOL.
This is straightforward for all expansion rules except for the last branching expansion rule.
To illustrate, suppose that we are to disprove a statement of the form \lstinline!s $\neq$ (t :: 'a) $\land$ s $\in$ (A :: 'a set) $\cup$ B $\land$ $\ldots$! in Isabelle/HOL
that corresponds to \lstinline!s' $\neqS$ t' $\fmAndSymbol$ s' $\inS$ A' $\unionS$ B' $\fmAndSymbol$ $\ldots$! in our set syntax.
Applying the decision procedure to this goal might result in a tableau proof containing an application of the last branching rule rule to \lstinline!(s' $\neqS$ t') $\in$ set b!, which creates the branches \lstinline![Var x $\inS$ s', Var x $\notinS$ t'] @ b! and \lstinline![Var x $\notinS$ s', Var x $\inS$ t'] @ b!.
But this can not be converted to a proof in Isabelle/HOL since \lstinline!s! and \lstinline!t! are not sets.

To deal with this problem, we introduce a light-weight type system that, given a mapping from variables to natural numbers, assigns a natural number to each set term.

\begin{itemize}
  \item Give example from HOL
  \item Show how branch is expanded
  \item Mention implemented constraint solver
  \item Show how definition of realise needs to be changed
  \item Syntax needs to be changed for empty set
\end{itemize}

@@

    Rules for \lstinline!Single! && Rules for \lstinline!$\eqS$! \\
    \cmidrule{1-1} \cmidrule{3-3}
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! \\
      \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! \\
      \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! \\
    \end{tabular}
    & &
    \begin{tabular}{p{0.22\textwidth}m{0.06\textwidth}p{0.2\textwidth}}
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\
      \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\
      \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\notinS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \end{tabular}
    \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules}
\end{table*}

@@

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b! \\
               \lstinline!$\land$! & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules. We write \lstinline!$\phi$! for \lstinline!last b! here.}
\end{table*}
